A Tale of Two Transects

### 2020-02-04 writing/transect_paper.Rmd
find strict consensus taxonomy for short amplicon sequences

Taxonomic assignments of ASVs which were represented in the long amplicon dataset were refined using the phylogenetic tree (Supplementary Figure \@ref(fig:phylotax)).
A taxon at a particular rank was assigned to a node and all its descendents if that taxon was consistent with the reference-based taxonomic assignments for each of the descendents.
A taxon assignment was considered to be consistent if at least one algorithm assigned that taxon at greater than 50% confidence, or if no algorithm successfully classified the sequence at greater than 50% confidence.
The result of this process was to give a taxonomic assignment to ASVs which were previously unassigned if they were nested within a clade where which was consistently given an assignment, as well as clarify the assignment of ASVs where different algorithms had resulted in different assignments, but only one of these was consistent with the assignments of other ASVs in the same clade.

@@

(ref:tax-chart) Fraction of ASVs (left) and reads (right) assigned to each taxonomic rank, for different sequencing technologies (PacBio RS II, Ion Torrent Ion S5), amplicons (Long, Short), reference databases (Unite, Warcup, RDP), and assignment algorithms (PHYLOTAX, RDPC, SINTAX, IDTAXA).

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
readd(taxon_reads, cache = cache) %>%
  pivot_longer(names_to = "rank", cols = c("kingdom", "phylum", "class", "order", "family", "genus"), values_to = "taxon") %>%
  group_by(amplicon, tech, Algorithm, reference, rank, ID = !is.na(taxon)) %>%
  summarize(reads = sum(reads), ASVs = n()) %>%
  group_by(amplicon, tech, Algorithm, reference, rank) %>%
  mutate(ASVs = ASVs / sum(ASVs)) %>%
  ungroup() %>%
  filter(ID) %>%
  mutate_at("rank", factor, levels = c("kingdom", "phylum", "class", "order", "family", "genus")) %>%
  rename(Reads = reads) %>%
  mutate_at("reference", as.character) %>%
  mutate(
    reference = map(reference, ~ if (. == "All") c("Unite", "Warcup", "RDP") else .)
  ) %>%
  unnest(reference) %>%
  mutate_at("reference", factor, levels = c("Unite", "Warcup", "RDP")) %>%
  pivot_longer(cols = c("Reads", "ASVs"), names_to = "type", values_to = "frac") %>%
  ggplot(aes(x = rank, y = frac, ymax = frac, group = Algorithm, fill = Algorithm)) +
  ggnomics::facet_nested(
    tech + amplicon ~ type + reference,
    resect = unit(1, "mm"),
    nest_line = TRUE,
    bleed = FALSE
  ) +
  # geom_ribbon(ymin = 0, alpha = 0.5) +
  # geom_line(aes(color = Algorithm)) +
  geom_col(position = "dodge") +
  ylab("Fraction of reads assigned") +
  xlab(NULL) +
  theme(axis.text.x = element_text(vjust = 0.5, angle = 90),
        legend.direction = "horizontal",
        legend.position = "bottom",
        strip.background = element_blank())
```

@@

light_plan %>%
  filter(startsWith(target, "correlog")) %>%
  select(target) %>%
  separate(
    target,
    into = c("step", "guild", "metric", "tech", "amplicon", "tdist"),
    sep = "_",
    remove = FALSE
  ) %>%
  mutate(
    tdist = paste(tdist, "year"),
    guild = plyr::mapvalues(guild, c("fungi", "ecm"), c("all Fungi", "ECM")),
    metric = plyr::mapvalues(metric, c("bray", "wunifrac"), c("Bray-Curtis", "W-UNIFRAC")),
    amplicon = factor(amplicon, levels = c("Short", "Long"))
  ) %>%
  mutate_at("target", purrr::map, readd, character_only = TRUE, cache = cache) %>%
  mutate_at("target", purrr::map, "mantel.res") %>%
  mutate_at("target", purrr::map, as.data.frame) %>%
  unnest(target) %>%
  filter(complete.cases(.)) %>%
  mutate(Significant = `Pr(corrected)` < 0.05) %>%
  ggplot(aes(class.index, Mantel.cor, group = tdist, color = tdist, shape = Significant)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.3, size = 1.5) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  # geom_line(aes(y = exp(predict(bc_fit)))) +
  scale_y_continuous(limits = c(-0.05, NA), name = "Mantel correlation") +
  scale_x_continuous(breaks = c(0, 3, 6, 9, 12)) +
  scale_color_discrete(name = "Time lag") +
  xlab("Distance (m)") +
  ggnomics::facet_nested(metric + tech + amplicon ~ guild, scales = "fixed", nest_line = TRUE) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1)) +
  theme(strip.background = element_blank())
```

@@

```{r spatio-temporal, fig.cap = "(ref:spatio)"}
variog_data <- light_plan %>%
  filter(step == "variofitST") %>%
  select(variofitST, variogST, guild, metric, tech, amplicon) %>%
  mutate_at(c("guild", "metric", "tech", "amplicon"), str_replace_all, "\"", "") %>%
  mutate(
    guild = plyr::mapvalues(guild, c("fungi", "ecm"), c("All Fungi", "ECM")),
    metric = plyr::mapvalues(metric, c("bray", "wunifrac"), c("Bray-Curtis", "W-UNIFRAC")),
    amplicon = factor(
      amplicon,
      levels = c("Short", "Long")
    )
  ) %>%
  mutate_at(
    c("variogST", "variofitST"),
    lapply,
    readd,
    character_only = TRUE,
    cache = cache)

@@

variog_empir <- variog_points %>%
  # mutate(bin = cut(dist, c(1:13 - 0.5, 14.5, 16.5, 19.5, 24.5))) %>%
  group_by(guild, metric, tech, amplicon, dist, timelag) %>%
  summarize(
    gamma = mean(gamma),
    # dist = mean(dist),
    n = n()
  )

variog_confint <- variog_points %>%
  group_by(guild, metric, tech, amplicon, dist, timelag) %>%
  summarize(
    lower = quantile(gamma, 0.025),
    upper = quantile(gamma, 0.975)
  )

@@

variog_empir %>%
  ggplot(aes(dist, gamma, color = timelag, size = n)) +
  # geom_hline(data = variog_sills, aes(yintercept = sills), linetype = "dashed", color = "gray50") +
  # geom_vline(data = variog_data, aes(xintercept = range), linetype = "dashed", color = "gray50") +
  # geom_jitter(shape = 46, alpha = 0.5) +
  # geom_line(size = 1.5, alpha = 0.5) +
  geom_point(alpha = 0.5) +
  # geom_smooth(data = variog_points, aes(fill = timelag), color = FALSE, alpha = 0.5) +
  geom_line(data = filter(variog_fit, timelag == 1), size = 0.5, color = "darkcyan", alpha = 0.8) +
  geom_line(data = filter(variog_fit, timelag == 0), size = 0.5, color = "red1", alpha = 0.8) +
  scale_x_continuous(expand = expand_scale(0, 0), limits = c(0, 25), name = "Distance (m)") +
  scale_color_discrete(name = "Time lag") +
  scale_size_continuous(name = "Comparisons") +
  scale_y_continuous(name = "Community dissimilarity") +
  ggnomics::facet_nested(metric + tech + amplicon ~ guild, scales = "free", nest_line = TRUE) +
  theme(strip.background = element_blank())

@@

```{r variofit}
variog_data %>%
  mutate_at("variofitST", lapply, purrr::compose(as.data.frame, as.list, gstat::extractPar)) %>%
  unnest(variofitST) %>%
  arrange(metric, amplicon, tech, guild) %>%
  select(Guild = guild, Amplicon = amplicon, Tech = tech, Metric = metric,
         Range = range, Anisotropy = anis) %>%
  mutate(Range = 3*Range,
         "Time Range" = Range/Anisotropy,
         "Time Range" = ifelse(`Time Range` > 100, NA, `Time Range`)) %>%
  mutate_at(c("Range", "Time Range"), formatC, digits = 1,
            format = "f") %>%
  mutate_at("Range", paste, "m") %>%
  mutate_at("Time Range", paste, "a") %>%
  mutate_at("Time Range", str_replace, "NA a", "-") %>%
  mutate_at("Amplicon", fct_relabel, str_replace, " amplicon", "") %>%
  select(Guild, Amplicon, Tech, Metric, "Space Range" = Range, "Time Range") %>%
  kable(booktabs = TRUE, caption = "(ref:variofit)")
```

@@

```{r short-its, fig.cap = ""}
readd(big_seq_table_ITS2, cache = cache) %>%
  colSums() %>%
  enframe(name = "ITS2", value = "nreads") %>%
  mutate(length = nchar(ITS2)) %>%
  filter(length <= 140) %>%
  mutate_at("ITS2", ~tzara::seqhash(chartr("T", "U", .))) %>%
  semi_join(readd(taxon_reads, cache = cache), ., by = c("label" = "ITS2")) %>%
  # group_by(ITS2) %>%
  # filter(!"Short" %in% amplicon) %>%
  taxon_plot(kingdom, y = reads)
```

>>>>

Taxonomic assignments of ASVs from the long amplicon dataset were refined using the phylogenetic tree (Supplementary Figure \@ref(fig:phylotax)).
A taxon at a particular rank was assigned to a node and all its descendents if that taxon was consistent with the reference-based taxonomic assignments for each of the descendents.
A taxon assignment was considered to be consistent if at least one algorithm assigned that taxon at greater than 50% confidence, or if no algorithm successfully classified the sequence at greater than 50% confidence.
The result of this process was to give a taxonomic assignment to ASVs which were previously unassigned if they were nested within a clade where which was consistently given an assignment, as well as clarify the assignment of ASVs where different algorithms had resulted in different assignments, but only one of these was consistent with the assignments of other ASVs in the same clade.
This refinement algorithm is referred to as "PHYLOTAX".

ASVs from the short amplicon datasets were refined using only the final strict consensus step, i.e., an assignment at a given rank was accepted if there was no conflict between the different assignment algorithms at greater than 50% confidence.
This refinement method is referred to as "Consensus".
Additionally, a hybrid method, "PHYLOTAX+C", was applied to the short amplicon datasets, in which assignments from PHYLOTAX were used for ASVs which could be linked by an identical ITS2 region to a long amplicon, and assignments from Consensus were used for the remaining ASVs.

@@

(ref:tax-chart) Fraction of ASVs (left) and reads (right) assigned to each taxonomic rank, for different sequencing technologies (PacBio RS II, Ion Torrent Ion S5), amplicons (Long, Short), reference databases (Unite, Warcup, RDP), and assignment algorithms (PHYLOTAX, PHYLOTAX+C, Consensus, RDPC, SINTAX, IDTAXA).

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
readd(taxon_reads, cache = cache) %>%
  pivot_longer(names_to = "rank", cols = c("kingdom", "phylum", "class", "order", "family", "genus"), values_to = "taxon") %>%
  group_by(amplicon, tech, Algorithm, reference, rank, ID = !is.na(taxon)) %>%
  summarize(reads = sum(reads), ASVs = n()) %>%
  group_by(amplicon, tech, Algorithm, reference, rank) %>%
  mutate(ASVs = ASVs / sum(ASVs)) %>%
  ungroup() %>%
  filter(ID) %>%
  mutate_at("rank", factor, levels = c("kingdom", "phylum", "class", "order", "family", "genus")) %>%
  rename(Reads = reads) %>%
  mutate_at("reference", as.character) %>%
  mutate_at("reference", replace_na, "All") %>%
  mutate(
    reference = map2(reference, amplicon, ~ if (.x == "All") c("Unite", "Warcup", if (.y == "Long") "RDP" else NULL) else .)
  ) %>%
  unnest(reference) %>%
  mutate_at("reference", factor, levels = c("Unite", "Warcup", "RDP")) %>%
  pivot_longer(cols = c("Reads", "ASVs"), names_to = "type", values_to = "frac") %>%
  ggplot(aes(x = rank, y = frac, ymax = frac, group = Algorithm, fill = Algorithm)) +
  ggnomics::facet_nested(
    tech + amplicon ~ type + reference,
    resect = unit(1, "mm"),
    nest_line = TRUE,
    bleed = FALSE
  ) +
  # geom_ribbon(ymin = 0, alpha = 0.5) +
  # geom_line(aes(color = Algorithm)) +
  geom_col(position = "dodge") +
  # scale_fill_brewer(type = "qual", palette = 2) +
  ylab("Fraction of reads assigned") +
  xlab(NULL) +
  theme(axis.text.x = element_text(vjust = 0.5, angle = 90),
        legend.direction = "horizontal",
        legend.position = "bottom",
        legend.
        strip.background = element_blank())
```

@@

light_plan %>%
  filter(startsWith(target, "correlog")) %>%
  select(target) %>%
  separate(
    target,
    into = c("step", "guild", "metric", "tech", "amplicon", "method", "tdist"),
    sep = "_",
    remove = FALSE
  ) %>%
  mutate(
    tdist = paste(tdist, "year"),
    guild = plyr::mapvalues(guild, c("fungi", "ecm"), c("all Fungi", "ECM")),
    metric = plyr::mapvalues(metric, c("bray", "wunifrac"), c("Bray-Curtis", "W-UNIFRAC")),
    amplicon = factor(amplicon, levels = c("Short", "Long"))
  ) %>%
  mutate_at("target", purrr::map, readd, character_only = TRUE, cache = cache) %>%
  mutate_at("target", purrr::map, "mantel.res") %>%
  mutate_at("target", purrr::map, as.data.frame) %>%
  unnest(target) %>%
  filter(complete.cases(.)) %>%
  mutate(Significant = `Pr(corrected)` < 0.05) %>%
  ggplot(aes(class.index, Mantel.cor, group = tdist, color = tdist, shape = Significant)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.3, size = 1.5) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  # geom_line(aes(y = exp(predict(bc_fit)))) +
  scale_y_continuous(limits = c(-0.05, NA), name = "Mantel correlation") +
  scale_x_continuous(breaks = c(0, 3, 6, 9, 12)) +
  scale_color_discrete(name = "Time lag") +
  xlab("Distance (m)") +
  ggnomics::facet_nested(metric + tech + amplicon + method ~ guild, scales = "fixed", nest_line = TRUE) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1)) +
  theme(strip.background = element_blank())
```

@@

```{r spatio-temporal, fig.cap = "(ref:spatio)"}
variog_data <- light_plan %>%
  filter(step == "variofitST") %>%
  select(variofitST, variogST, guild, metric, tech, amplicon, algorithm) %>%
  mutate_at(c("guild", "metric", "tech", "amplicon", "algorithm"), str_replace_all, "\"", "") %>%
  mutate(
    guild = plyr::mapvalues(guild, c("fungi", "ecm"), c("All Fungi", "ECM")),
    metric = plyr::mapvalues(metric, c("bray", "wunifrac"), c("Bray-Curtis", "W-UNIFRAC")),
    amplicon = factor(
      amplicon,
      levels = c("Short", "Long")
    ),
    algorithm = factor(
      algorithm,
      levels = c("Consensus", "PHYLOTAX", "PHYLOTAX+Cons"),
      labels = c("Cons", "PHYLO", "PHYLO+C")
    )
  ) %>%
  mutate_at(
    c("variogST", "variofitST"),
    lapply,
    readd,
    character_only = TRUE,
    cache = cache)

@@

variog_empir <- variog_points %>%
  # mutate(bin = cut(dist, c(1:13 - 0.5, 14.5, 16.5, 19.5, 24.5))) %>%
  group_by(guild, metric, tech, amplicon, algorithm, dist, timelag) %>%
  summarize(
    gamma = mean(gamma),
    # dist = mean(dist),
    n = n()
  )

variog_confint <- variog_points %>%
  group_by(guild, metric, tech, amplicon, algorithm, dist, timelag) %>%
  summarize(
    lower = quantile(gamma, 0.025),
    upper = quantile(gamma, 0.975)
  )

@@

variog_empir %>%
  ggplot(aes(dist, gamma, color = timelag, size = n)) +
  # geom_hline(data = variog_sills, aes(yintercept = sills), linetype = "dashed", color = "gray50") +
  # geom_vline(data = variog_data, aes(xintercept = range), linetype = "dashed", color = "gray50") +
  # geom_jitter(shape = 46, alpha = 0.5) +
  # geom_line(size = 1.5, alpha = 0.5) +
  geom_point(alpha = 0.5) +
  # geom_smooth(data = variog_points, aes(fill = timelag), color = FALSE, alpha = 0.5) +
  geom_line(data = filter(variog_fit, timelag == 1), size = 0.5, color = "darkcyan", alpha = 0.8) +
  geom_line(data = filter(variog_fit, timelag == 0), size = 0.5, color = "red1", alpha = 0.8) +
  scale_x_continuous(expand = expand_scale(0, 0), limits = c(0, 25), name = "Distance (m)") +
  scale_color_discrete(name = "Time lag") +
  scale_size_continuous(name = "Comparisons") +
  scale_y_continuous(name = "Community dissimilarity") +
  ggnomics::facet_nested(metric + tech + amplicon + algorithm ~ guild, scales = "free", nest_line = TRUE) +
  theme(strip.background = element_blank())

@@

```{r variofit}
variog_data %>%
  mutate_at("variofitST", lapply, purrr::compose(as.data.frame, as.list, gstat::extractPar)) %>%
  unnest(variofitST) %>%
  arrange(metric, amplicon, tech, guild) %>%
  select(Amplicon = amplicon, Tech = tech, Guild = guild, Algorithm = algorithm,
         Metric = metric, Range = range, Anisotropy = anis) %>%
  mutate(Range = 3*Range,
         "Time Range" = Range/Anisotropy,
         "Time Range" = ifelse(`Time Range` > 100, NA, `Time Range`)) %>%
  mutate_at(c("Range", "Time Range"), formatC, digits = 1,
            format = "f") %>%
  mutate_at("Range", paste, "m") %>%
  mutate_at("Time Range", paste, "a") %>%
  mutate_at("Time Range", str_replace, "NA a", "-") %>%
  mutate_at("Amplicon", fct_relabel, str_replace, " amplicon", "") %>%
  select(Amplicon, Tech, Guild, Algorithm, Metric, "Space Range" = Range, "Time Range") %>%
  kable(booktabs = TRUE, caption = "(ref:variofit)")
```

@@

```{r short-its, fig.cap = "(ref:short-its)"}
readd(big_seq_table_ITS2, cache = cache) %>%
  colSums() %>%
  enframe(name = "ITS2", value = "nreads") %>%
  mutate(length = nchar(ITS2)) %>%
  filter(length <= 140) %>%
  mutate_at("ITS2", ~tzara::seqhash(chartr("T", "U", .))) %>%
  semi_join(readd(taxon_reads, cache = cache), ., by = c("label" = "ITS2")) %>%
  # group_by(ITS2) %>%
  # filter(!"Short" %in% amplicon) %>%
  taxon_plot(kingdom, y = reads)
```
